---
title: "Comparing Sentiment Analysis Models in R"
author: "Alex Crest & Aybuke Atalay"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

Hello World! I hope you are all ready to get stuck in with sentiment.
(positive: 0.071, neutral: 0.857, negative: 0) Today, we are looking to
build our knowledge, and to (positive: 0, neutral: 1, negative: 0)
provide you with the tools to conduct sentiment analysis in your own
research. (positive: 0, neutral: 1, negative: 0)

This is the MD file for Comparing Sentiment Analysis in R. It is
designed to be used as part of a workshop, along with a presentation.
Please refer to the associated presentation for introduction and initial
guidance.

Before the session starts, make sure you have R 4.5.3 installed and
running!

```{r}
R.version.string
install.packages("installr")
library(installr)
#updateR()
```

# Schedule

Session start 14:05

14:05 - 14:10 1) Housekeeping

14:10 - 14:20 2) Introduction to Sentiment Analysis: What, Why, How?

14:20 - 15:50 3) A Comparative look at Three Examples (with a 10 minute
break)

15:50 - 16:00 4) Final Q and A.

## 1) Housekeeping and Set Up

### Aims

Understanding what sentiment analysis is and its utility Processing data
Understanding three packages: tidytext, syuzhet, quanteda Building
skills in R Summarising and comparing sentiment results with other
variables Saving data including to plots, .csv, and .json.

### Setting

The work will be completed between this markdown and your own local
environment on R.

Open RStudio

Go to File → New Project → Version Control → Git

Paste the repo URL from teams

Choose a folder

Click Create Project

We're ready to go!

## 2) Introduction to Sentiment Analysis: What, Why, How?

Focus on the presentation, nothing needed here just now!

## 3) Practical: A Comparative look at Three Examples

Before any text analysis, your data needs to be preprocessed. Today we
will be working with three datasets: A collection of book reviews
from Amazon (Courtesy of Hou et al. 2024). The Gettysberg Address,
courtesy of Abraham Lincoln. The inaugural addresses from US presidents
from 1900, already present in the quanteda package.

We will be using three different tools to assess these: sentimentr -
Amazon Reviews, lots of short texts syuzhet - the Gettysburg Address, a
single narrative quanteda - Inaugural Adresses, political speeches with
multiple subjects

Using the `Tidyverse` functions, we have a variety of very useful tools
to tidy the data, and some that are essential before sentiment analysis
can begin. From within this collection we are mostly going to use `Purrr`, `readr`, and `ggplot`.

We will be using the following packages for today

Some of these are already installed in Noteable so run the cell below
only if you are working locally and you have never used any of these
packages before

```{r}
{install.packages(c("tidyverse", "sentimentr", "jsonlite", "syuzhet", "quanteda","tokenizers", "textdata"))}
```

Run this if you are working on Noteable

```{r}
install.packages(c("sentimentr","syuzhet","textdata"))
```

```{r}
library(tidyverse)
library(sentimentr)
library(syuzhet)
library(jsonlite)
library(quanteda)
library(textdata)
library(tokenizers)
```

If you have any issues installing these, let us know now, and we can fix
the issue promptly.

## Accessing the data

Now we have our environment, we need to load the data! Now we need to
load this data into our workspace. Usually, we would have to find it,
either from an online repository, collect or build it ourselves.
However, today, the Amazon Reviews and The Gettsberg Address are already
loaded into the directory. We can also load the presidential addresses.
When you loaded the git, two datasets should have come with the rest of
the files. The Amazon dataset is called 'Data_Books_Lite.json' and
contains reviews from Amazon on different books.

```{r}
Amazon_Reviews <- fromJSON("Data_Books_Lite.json")
```

The Gettysburg address can be found here:

```{r}
Gettysburg_Address <- tibble(text = read_lines("Gettysburg.txt"))
```

The inaugural addresses are a dataset that already exists in quanteda!
We just need to select the relavent data.

```{r}
Inaugural_Address_Post_1900 <- corpus_subset(
  data_corpus_inaugural, Year >= 1900)
```

Now convert the object to a dataframe with associated metadata

```{r}
Inaugural_Addresses <- data.frame(
  Year = docvars(Inaugural_Address_Post_1900, "Year"),
  President = docvars(Inaugural_Address_Post_1900, "President"),
  Text = as.character(Inaugural_Address_Post_1900)
)
```

You will notice that each medium has different characteristics (long
speech, short phrases, punctuation etc.) that need to be considered for
the type of sentiment analysis.

It is best practice to also create a backup of your data in case any
step goes wrong.

``` {r}        
Backup_Inaugural_Addresses <- Inaugural_Addresses
```

```{r}
Backup_Amazon_Reviews <- Amazon_Reviews
```

Create backups of the other two datasets and run these locally

To reload the backup, simply reverse it

```{r}
Amazon_Reviews <- Backup_Amazon_Reviews
```

There are three steps to sentiment Analysis: Text Preprocessing,
Applying Analysis, Interpreting the Results.

Thankfully, all of our datasets are structured as dataframes and our
preprocesing is going to be text specific. They are also clean, so this
is a relatively painless process. We need to transform raw text into a
more consistent and analysable form before applying sentiment analysis

We are going to go through steps 1 - 3 for each dataset on its own, and
then compare in more detail.

# Amazon Reviews

This is a big dataset, larger than a single researcher could reasonably
quantitively review. Note that the methods today can be scaled up for
larger datasets still!

First, let's **inspect** the data to understand its composition

```{r}
head(Amazon_Reviews)
colnames(Amazon_Reviews)
sapply(Amazon_Reviews, class)
```

We will keep all the columns for now, but they provide useful variables
for comparison later on. The column we are concerned with for sentiment
is 'text', containing the raw text data.

In the following line we will **preprocess** the data: Reduce all text
to lowercase, and remove extra spaces. We are going to use 'mutate',
which creates a new column, from all the instructions after the pipe
'%\>%'

```{r}
Amazon_Reviews <- Amazon_Reviews %>%
  mutate(
    Text_Processed = text %>%
      str_to_lower() %>%
      str_squish()
  )
```

Now we calculate the sentiment score column.

```{r}
review_sentiment <- sentiment_by(Amazon_Reviews$Text_Processed)
```

sentimentr calculates sentiment by sentence automatically, and then
combines these scores into an average for the document.

We need to add those averages back into our data:

```{r}
Amazon_Reviews <- Amazon_Reviews %>%
  mutate(
    sentimentr_score = review_sentiment$ave_sentiment,
    sentiment_label = case_when(
      sentimentr_score > 0 ~ "positive",
      sentimentr_score < 0 ~ "negative",
      TRUE ~ "neutral"
    )
  )
```

You have results for each row! This produces a sentiment score for each
review, along with a categorical label indicating whether the review is
positive, negative, or neutral.

To begin **interpretting** the data, find the mean across all reviews:

```{r}
sentiment_mean <- Amazon_Reviews %>%
  summarise(
    mean_sentiment = mean(sentimentr_score, na.rm = TRUE)
  )
```

```{r}
print(sentiment_mean)
```

This indicates the overall polarity of the review corpus. Values above
zero suggest a generally positive set of reviews, while values below
zero suggest a more negative corpus.

We can now calculate the average sentiment for each book by grouping
reviews according to `meta_title`.

```{r}
book_sentiment <- Amazon_Reviews %>%
  group_by(meta_title) %>%
  summarise(
    mean_net_sentiment = mean(sentimentr_score, na.rm = TRUE),
    review_count = n()
  ) %>%
  ungroup()
```

You can manually sift through per book. This dataset is already made of
precisely 10 reviews per book.

Now, we are going to investigate the extremities of our data: the books
with the highest and lowest sentiment

```{r}
top_3 <- book_sentiment %>%
  arrange(desc(mean_net_sentiment)) %>%
  slice_head(n = 3)

bottom_3 <- book_sentiment %>%
  arrange(mean_net_sentiment) %>%
  slice_head(n = 3)

plot_data <- bind_rows(top_3, bottom_3)
```

If you have worked in R before, you may have used ggplot or ggplot2 to
plot your data. Sentiment can be plotted much as any other variable, but
here we will guide you through best practices and appropriate plots for
the models we have used.

Below, we will create a bar chart of the sentiment scores

```{r}
ggplot(plot_data, 
       aes(y = reorder(str_c(word(meta_title, 1, 3), "..."), 
                       mean_net_sentiment),
           x = mean_net_sentiment)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Extreme Net Sentiment by Book",
    x = "Mean Net Sentiment",
    y = "Book"
  )
```

Now **save and export** the graph

```{r}
ggsave("Extreme_Net_Sentiment_by_Book.png")
```

# The Gettysberg Address

This is a much simpler operation, as we have one row and one object that
we are going to analyse. This time we are going to devide the text up by
sentence, to track sentiment as it develops through Lincoln's words

```{r}
sentences <- get_sentences(Gettysburg_Address$text)
sentiment_values <- get_sentiment(sentences, method = "syuzhet")
sentiment_df <- data.frame(
  sentence = sentences,
  sentiment = sentiment_values,
  index = 1:length(sentences)
)
```

Now the trajectory of sentiment through the text can be plotted

```{r}
ggplot(sentiment_df, aes(x = index, y = sentiment)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Sentiment Trajectory of the Gettysburg Address",
    x = "Sentence Number",
    y = "Sentiment"
  ) +
  theme_minimal()
  
ggsave("gettysburg_sentiment_plot.png")
```

It is so much simpler! More so, this can be combined with other analysis
and compared with other speeches! We can assess sentiment by theme, time
in the speech and other qualitative variables also.

#Inaugural Presidential Adrresses For political speeches with multiple
subjects, we will be using the quanteda package. This time, our analysis
is going to be more complex.

We need to convert the texts into a corpus first, so our package can
handle it

```{r}
corp <- corpus(Inaugural_Addresses, text_field = "Text")
```

Then tokenise, and remove extraneous data and create a document feature
matrix

```{r}
toks <- tokens(
  corp,
  remove_punct = TRUE,
  remove_numbers = TRUE,
  remove_symbols = TRUE
)
dfmat <- dfm(toks)
```

Count total words in each speech

```{r}
word_counts <- ntoken(toks)
```

Apply the built in LSD2015 dictionary to the data

```{r}
sent_dfm <- dfm_lookup(dfmat, dictionary = data_dictionary_LSD2015)
```

## Understanding polarity

Option 1. Raw Sentiment

```{r}
sentiment_results <- convert(sent_dfm, to = "data.frame") %>%
  mutate(
    Year = docvars(corp, "Year"),
    President = docvars(corp, "President"),
    net_sentiment = positive - negative
  )
```

Option 2. Normalised

```{r}
sentiment_results <- convert(sent_dfm, to = "data.frame") %>%
  mutate(
    Year = docvars(corp, "Year"),
    President = docvars(corp, "President"),
    word_count = word_counts,
    net_sentiment = positive - negative,
    normalized_sentiment = net_sentiment / word_count
  )
```

These scores have been **normalised** by dividing them by the total
tokens in each speech/document. This is a crucial step to providing
comparable data across a corpus. Otherwise, the presidents that spoke
more would simply have higher scores. this is itself a finding

```{r}
ggplot(sentiment_results,
       aes(x = Year, y = normalized_sentiment)) +
  geom_col() +
  theme_minimal() +
  labs(
    title = "Normalised Net Sentiment in Post-1900 Inaugural Addresses",
    x = "Year",
    y = "Net Sentiment per Word"
  )
```

# Comparison

We now have three different datasets, analysed in three slightly
different ways by virtue of the different packages we applied. Each has
different results. Now we are going to return to our Amazon reviews and
compare the results as they work on just one dataset.

## Run quanteda

Develop the Corpus

```{r}
corp_amazon <- corpus(Amazon_Reviews, text_field = "Text_Processed")
```

Great, now tokenize.

```{r}
toks_amazon <- tokens(
  corp_amazon,
  remove_punct = TRUE,
  remove_numbers = TRUE,
  remove_symbols = TRUE
)
dfmat_amazon <- dfm(toks_amazon)

```

Run the analysis, and combine with the original dataset

```{r}
sent_dfm_amazon <- dfm_lookup(dfmat_amazon, dictionary = data_dictionary_LSD2015)

sentiment_quanteda <- convert(sent_dfm_amazon, to = "data.frame") %>%
  mutate(
    net_sentiment_q = positive - negative,
    normalised_sentiment_q = net_sentiment_q / ntoken(toks_amazon)
  )
Amazon_Reviews <- bind_cols(
  Amazon_Reviews,
  sentiment_quanteda %>% select(net_sentiment_q, normalised_sentiment_q)
)
```

## Calculate the correlation between the two scores

```{r}
cor(
  Amazon_Reviews$sentimentr_score,
  Amazon_Reviews$normalised_sentiment_q,
  use = "complete.obs"
)
```

What is the utility of doing this?

## Visualise them

```{r}
ggplot(Amazon_Reviews,
       aes(x = sentimentr_score, y = normalised_sentiment_q)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  labs(
    title = "Comparison of Sentiment Methods",
    x = "Sentimentr Score",
    y = "Quanteda Normalized Sentiment"
  )
```

## Apply syuzhet to complete the triad

```{r}
Amazon_Reviews <- Amazon_Reviews %>%
  mutate(
    syuzhet_score = get_sentiment(Text_Processed, method = "syuzhet")
  )
```

Visualise the results

```{r}
ggplot(Amazon_Reviews) +
  geom_density(aes(x = sentimentr_score, linetype = "sentimentr")) +
  geom_density(aes(x = normalised_sentiment_q, linetype = "quanteda")) +
  geom_density(aes(x = syuzhet_score, linetype = "syuzhet")) +
  coord_cartesian(xlim = c(-1, 1)) +
  labs(
    title = "Distribution of Sentiment Scores Across Methods",
    x = "Sentiment Score",
    y = "Density",
    linetype = "Method"
  ) +
  theme_minimal()
```

## Compare them all

```{r}
comparison_books <- Amazon_Reviews %>%
  group_by(meta_title) %>%
  summarise(
    sentimentr_mean = mean(sentimentr_score, na.rm = TRUE),
    quanteda_mean = mean(normalised_sentiment_q, na.rm = TRUE),
    syuzhet_mean = mean(syuzhet_score, na.rm = TRUE),
    review_count = n()
  ) %>%
  ungroup()
```

You can inspect it with:

```{r}
print(comparison_books)
```

Or plot one method against another:

```{r}
ggplot(comparison_books, aes(x = sentimentr_mean, y = quanteda_mean)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Book-Level Sentiment: Sentimentr vs Quanteda",
    x = "Mean Sentimentr Score",
    y = "Mean Quanteda Score"
  ) +
  theme_minimal()
```

And similarly for syuzhet:

```{r}
ggplot(comparison_books, aes(x = sentimentr_mean, y = syuzhet_mean)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Book-Level Sentiment: Sentimentr vs Syuzhet",
    x = "Mean Sentimentr Score",
    y = "Mean Syuzhet Score"
  ) +
  theme_minimal()
```

### Compare rankings of books across methods

```{r}
top_3_sentimentr <- comparison_books %>%
  arrange(desc(sentimentr_mean)) %>%
  slice_head(n = 3)
bottom_3_sentimentr <- comparison_books %>%
  arrange(sentimentr_mean) %>%
  slice_head(n = 3)
```

```{r}
top_3_quanteda <- comparison_books %>%
  arrange(desc(quanteda_mean)) %>%
  slice_head(n = 3)
bottom_3_quanteda <- comparison_books %>%
  arrange(quanteda_mean) %>%
  slice_head(n = 3)
```

```{r}
top_3_syuzhet <- comparison_books %>%
  arrange(desc(syuzhet_mean)) %>%
  slice_head(n = 3)
bottom_3_syuzhet <- comparison_books %>%
  arrange(syuzhet_mean) %>%
  slice_head(n = 3)
```

```{r}
book_rankings <- comparison_books %>%
  mutate(
    rank_sentimentr = rank(-sentimentr_mean, ties.method = "first"),
    rank_quanteda = rank(-quanteda_mean, ties.method = "first"),
    rank_syuzhet = rank(-syuzhet_mean, ties.method = "first")
  ) %>%
  select(meta_title, rank_sentimentr, rank_quanteda, rank_syuzhet)

print(book_rankings)
```

## Optional: We can also use sentiment analysis to model emotions in text

## If instructed, search for 'End of Session'
Tidytext is already installed in noteable so only run the below if you are running this locally otherwise just go directly to the next cell

```{r}         
install.packages(tidytext)
```

```{r}
library(tidytext)
```

In tidytext, there is a sentiment analysis lexicon called NRC, a
dictionary of thousands of terms that have been tagged positive,
negative, and also given an emotional tag: anger fear joy sadness trust
disgust surprise anticipation

below is a method for processing said analysis

```{r}
NRC_Amazon_Reviews <- Amazon_Reviews %>%
  mutate(Text_Processed = text %>%                    # creates a new column
           str_to_lower() %>%                         # lowercase
           # str_replace_all("[^a-z\\s]", "") %>%     # Optional, remove punctuation/numbers. Some sentiment tools struggle with punctuation. Today, we will remove it. We can always repeat and omit this line
           str_squish()                               # remove extra spaces and line breaks
  )

```

We have created a new column, and preserved the raw data for the method
we are using, we need to tokenise, that is devide up the sentences into
lists of words

```{r}
NRC_Amazon_Reviews <- NRC_Amazon_Reviews %>%
  mutate(tokens = tokenize_words(Text_Processed))
```

A note of caution, text preprocessing often includes removing stopwords,
but words such as 'no' 'yes' 'he' 'here' are meaning-laden and their
removal will affect sentiment scores. We will leave them in, for now.

If you have any issues, reload the backup. By now, you should have a
dataframe with an additional column of "Text_Processed".

```{r}
colnames(NRC_Amazon_Reviews)
```

Load the lexicon

```{r}
nrc <- get_sentiments("nrc")
```

Now create a reference list of NRC words grouped by sentiment category

```{r}
nrc_lookup <- split(nrc$word, nrc$sentiment)
```

Apply it, and assign scores to each review We are also going to nest the
results in each row in a new column called sentiment_scores, for
tidiness

```{r}
NRC_Amazon_Reviews <- NRC_Amazon_Reviews %>%                             # #start with reviews dataset that already contains token lists
  mutate(
    sentiment_scores = purrr::map(tokens, \(x) {                         # #for each review, work through its tokens
      tibble::as_tibble_row(
        sapply(nrc_lookup, \(words) sum(x %in% words))                   # #count how many tokens match each NRC sentiment category
      )
    })
  )    
```

To make sense of this, unnest the sentiment scores.

```{r}
NRC_Amazon_Reviews <- NRC_Amazon_Reviews %>%
  unnest(sentiment_scores) %>%
  rename(
    nrc_anger = anger,
    nrc_anticipation = anticipation,
    nrc_disgust = disgust,
    nrc_fear = fear,
    nrc_joy = joy,
    nrc_sadness = sadness,
    nrc_surprise = surprise,
    nrc_trust = trust,
    nrc_positive = positive,
    nrc_negative = negative
  )
```

For ease of processing, we are going to convert the negative sentiment
to negative numbers

```{r}
NRC_Amazon_Reviews <- NRC_Amazon_Reviews %>%
  mutate(
    nrc_negative = -nrc_negative
  )
```

You can now aggregate these emotions according to different variables

## End of Session

This workshop applied three sentiment analysis approaches to different
textual datasets. `sentimentr` provided context-sensitive polarity
scoring for Amazon reviews, `syuzhet` enabled sentence-level tracking
across the Gettysburg Address, and `quanteda` supported scalable
dictionary-based analysis of inaugural addresses. Comparing these
methods showed that sentiment results vary depending on the package and
level of analysis, highlighting the importance of choosing a method
appropriate to the research question.

We only have one session today, but if you would like more tailored
feedback or assistance, please book a data surgery with us.
<https://www.cdcs.ed.ac.uk/training/data-surgery>

# Further Reading

For more on the utility of sentiment analysis and their critiques, read
the following papers.

Devika, M. D., Sunitha, C., & Ganesh, A. (2016). Sentiment analysis: a
comparative study on different approaches. Procedia Computer Science,
87, 44-49.

Gonçalves, P., Araújo, M., Benevenuto, F., & Cha, M. (2013, October).
Comparing and combining sentiment analysis methods. In Proceedings of
the first ACM conference on Online social networks (pp. 27-38).

Shiha, M., & Ayvaz, S. (2017). The effects of emoji in sentiment
analysis. Int. J. Comput. Electr. Eng.(IJCEE.), 9(1), 360-369.

Wankhade, M., Rao, A. C. S., & Kulkarni, C. (2022). A survey on
sentiment analysis methods, applications, and challenges. Artificial
intelligence review, 55(7), 5731-5780.


# A Tutorial for Advanced Sentiment in Python

<https://github.com/DCS-training/SentimentAnalysis>

# References:

Hou, Y., Li, J., He, Z., Yan, A., Chen, X., & McAuley, J. (2024).
Bridging language and items for retrieval and recommendation. arXiv
Preprint arXiv:2403.03952.

Saif M. Mohammad, S. M., & Peter D. Turney, P. D. (2013).Crowdsourcing a
Word–Emotion Association Lexicon.Computational Intelligence, 29(3),
436–465
